\vspace*{-40pt}Numerical Eigenvalue Algorithms

### 2025-03-06 documents/paper.tex
unlimited power
benchmarks

\usepackage{enumitem}
\usepackage{amsmath, amsfonts, amsthm, amssymb}
\usepackage{graphicx, float}
\usepackage[style=alphabetic]{biblatex}
\usepackage{mathtools}
\usepackage{mlmodern}
\usepackage{titlesec}
\usepackage{interval}
\usepackage{hyperref}
\usepackage{siunitx}
\usepackage{titling}
\usepackage{vwcol}
\usepackage{setspace}
\usepackage{empheq}
\usepackage{tabularray}
\usepackage{cancel}
\usepackage{esdiff}
\usepackage{multicol}
\usepackage{fancyvrb}
\usepackage{mdframed}
\usepackage{esdiff}
\usepackage{tikzsymbols}
\usepackage{multicol}
\usepackage{tikz}
\usepackage{varwidth}
\usepackage{parskip}
\usepackage{pgfplots}
\usepackage{soul}
\usepackage{tcolorbox}
\pgfplotsset{compat=1.18}
\addbibresource{refs.bib}
\intervalconfig {
	soft open fences
}

@@

\title{\vspace*{-40pt}Numerical Eigenvalue Algorithms}
\author{Jayden Li}
\date{Linear Algebra 2024-25}
% \allowdisplaybreaks
\postdisplaypenalty=100000
\setcounter{section}{-1}

@@

@@

Let $v_0\in\mathbb R^n$ be a random vector, let $v=Pv_0$, let $v_{0_i}$ be the $i$th component of $v_0$, and let $p_i$ be the $i$th column of $P$. Since $P$ is invertible, $v$ is essentially another random vector. $p_i$ must be an eigenvector of $A$, since $P$ is the matrix of eigenvectors. Multiplying on the left of equation (1), we have:
\begin{align*}
	A^kPv_0=PD^kv_0
	=A^k v 
	&= P \begin{bmatrix}
		\lambda_1^k & & \\
		& \ddots & \\
		& & \lambda_n^k
	\end{bmatrix}
	\begin{bmatrix}
		v_{0_1} \\ \dots \\ v_{0_n}
	\end{bmatrix}
	=\begin{bmatrix}
		p_1 & \ldots & p_n
	\end{bmatrix} \begin{bmatrix}
		\lambda_1^k v_{0_1} \\ 
		\vdots \\
		\lambda_n^k v_{0_n} \\ 
	\end{bmatrix}
	=\sum_{i=1}^{n}\lambda_i^k v_{0_i} p_i \\
	&=\sum_{i=1}^{n}\lambda_1^k\frac{\lambda_i^k}{\lambda_1^k} v_{0_i} p_i
	=\lambda_1^k \sum_{i=1}^{n}\left(\frac{\lambda_i}{\lambda_1}\right)^k v_{0_i}p_i
\end{align*}
\ul{Suppose the dominant eigenvalue $\lambda_1$ is greater than all other eigenvalues: $\left|\lambda_1\right|>\left|\lambda_2\right|\geq \left|\lambda_3\right|\geq\ldots\geq \left|\lambda_n\right|$}*, so $0\leq\lambda_i/\lambda_1<1$ for all $1\leq i\leq n$. Thus, as $k\to\infty$, $\left(\lambda_i/\lambda_1\right)^k\to0$.
\begin{equation*}
    \lim_{k\to\infty}A^kv
	=\lim_{k\to\infty}\lambda_1^k \sum_{i=1}^{n}\left(\frac{\lambda_i}{\lambda_1}\right)^k v_{0_i}p_i
	=\lim_{k\to\infty}\lambda_1^k \left(\cancel{\left(\frac{\lambda_1}{\lambda_1}\right)^k} v_{0_1}p_1+\cancel{\sum_{i=2}^{n}\left(\frac{\lambda_i}{\lambda_1}\right)^k v_{0_i}p_i}\right)
	=\lim_{k\to\infty}\lambda_1^k v_{0_1}{p_1}
\end{equation*}
Therefore, as $k$ tends to infinity, $A^kv$ is parallel to $p_1$ as $\lambda_1$ and $v_{0_1}$ are scalar values. Recall that $p_1$ is an eigenvalue of $A$, associated with the eigenvalue $\lambda_1$.

@@

Let us return to the condition marked by an asterisk above. If that assumption is not met, i.e. $\left|\lambda_1\right|=\left|\lambda_2\right|$. Let us suppose for now that $\left|\lambda_2\right|>\left|\lambda_3\right|$.
\begin{align*}
    \lim_{k\to\infty}A^k v
	&=\lim_{k\to\infty}\lambda_1^k \sum_{i=1}^{n}\left(\frac{\lambda_i}{\lambda_1}\right)^k v_{0_i}p_i
	=\lim_{k\to\infty}\lambda_1^k \left(\cancel{\left(\frac{\lambda_1}{\lambda_1}\right)^k} v_{0_1}p_1+\cancel{\left(\frac{\lambda_2}{\lambda_1}\right)^k}v_{0_2}p_2+\cancel{\sum_{i=3}^{n}\left(\frac{\lambda_i}{\lambda_1}\right)^k v_{0_i}p_i}\right) \\
	&=\lim_{k\to\infty}\left(v_{0_1}p_1+v_{0_2}p_2\right)
\end{align*}
Knowing this is parallel to $A^kv$ is not helpful, so the algorithm will break down.

@@

>>>>

\usepackage{enumitem}
\usepackage{amsmath, amsfonts, amsthm, amssymb}
\usepackage{graphicx, float}
\usepackage[style=alphabetic]{biblatex}
\usepackage{mathtools}
\usepackage{titlesec}
\usepackage{interval}
\usepackage{hyperref}
\usepackage{siunitx}
\usepackage{titling}
\usepackage{vwcol}
\usepackage{setspace}
\usepackage{empheq}
\usepackage{tabularray}
\usepackage{cancel}
\usepackage{esdiff}
\usepackage{multicol}
\usepackage{fancyvrb}
\usepackage{mdframed}
\usepackage{esdiff}
\usepackage{tikzsymbols}
\usepackage{multicol}
\usepackage{tikz}
\usepackage{varwidth}
\usepackage{parskip}
\usepackage{pgfplots}
\usepackage{soul}
\usepackage{tcolorbox}
\usepackage{fontspec}

\pgfplotsset{compat=1.18}
\addbibresource{refs.bib}
\intervalconfig {
	soft open fences
}

@@

\title{\vspace*{-40pt}Numerical Eigenvalue Algorithms}
\author{Jayden Li}
\date{Linear Algebra 2024-25}
% \allowdisplaybreaks
% \postdisplaypenalty=100000
\setcounter{section}{-1}

@@

\def\mathdefault#1{#1}
\everymath=\expandafter{\the\everymath\displaystyle}

@@

Let $v_0\in\mathbb R^n$ be a random vector, let $v=Pv_0$, let $v_{0_i}$ be the $i$th component of $v_0$, and let $p_i$ be the $i$th column of $P$. Since $P$ is invertible, $v$ is essentially another random vector. $p_i$ must be an eigenvector of $A$, since $P$ is the matrix of eigenvectors. Multiplying on the left of equation (1), we have:
\begin{align*}
	A^kPv_0=PD^kv_0
	=A^k v 
	&= P \begin{bmatrix}
		\lambda_1^k & & \\
		& \ddots & \\
		& & \lambda_n^k
	\end{bmatrix}
	\begin{bmatrix}
		v_{0_1} \\ \vdots \\ v_{0_n}
	\end{bmatrix}
	=\begin{bmatrix}
		p_1 & \ldots & p_n
	\end{bmatrix} \begin{bmatrix}
		\lambda_1^k v_{0_1} \\ 
		\vdots \\
		\lambda_n^k v_{0_n} \\ 
	\end{bmatrix}
	=\sum_{i=1}^{n}\lambda_i^k v_{0_i} p_i \\
	&=\sum_{i=1}^{n}\lambda_1^k\frac{\lambda_i^k}{\lambda_1^k} v_{0_i} p_i
	=\lambda_1^k \sum_{i=1}^{n}\left(\frac{\lambda_i}{\lambda_1}\right)^k v_{0_i}p_i
\end{align*}
\ul{Suppose the dominant eigenvalue $\lambda_1$ is greater than all other eigenvalues: $\left|\lambda_1\right|>\left|\lambda_2\right|\geq \left|\lambda_3\right|\geq\ldots\geq \left|\lambda_n\right|$}*, so $0\leq\lambda_i/\lambda_1<1$ for all $1\leq i\leq n$. Thus, as $k\to\infty$, $\left(\lambda_i/\lambda_1\right)^k\to0$.
\begin{equation*}
    \lim_{k\to\infty}A^kv
	=\lim_{k\to\infty}\lambda_1^k \sum_{i=1}^{n}\left(\frac{\lambda_i}{\lambda_1}\right)^k v_{0_i}p_i
	=\lim_{k\to\infty}\lambda_1^k \left(\cancel{\left(\frac{\lambda_1}{\lambda_1}\right)^k} v_{0_1}p_1+\cancel{\sum_{i=2}^{n}\left(\frac{\lambda_i}{\lambda_1}\right)^k v_{0_i}p_i}\right)
	=\lim_{k\to\infty}\lambda_1^k v_{0_1}{p_1}
\end{equation*}
Therefore, as $k$ tends to infinity, $A^kv$ is parallel to $p_1$ as $\lambda_1$ and $v_{0_1}$ are scalar values. Recall that $p_1$ is an eigenvalue of $A$, associated with the eigenvalue $\lambda_1$.

@@

Let us return to the condition marked by an asterisk above. If that assumption is not met, i.e. $\left|\lambda_1\right|=\left|\lambda_2\right|$. Let us suppose for now that $\left|\lambda_2\right|>\left|\lambda_3\right|$.
\begin{align*}
    \lim_{k\to\infty}A^k v
	&=\lim_{k\to\infty}\lambda_1^k \sum_{i=1}^{n}\left(\frac{\lambda_i}{\lambda_1}\right)^k v_{0_i}p_i
	=\lim_{k\to\infty}\lambda_1^k \left(\cancel{\left(\frac{\lambda_1}{\lambda_1}\right)^k} v_{0_1}p_1+\cancel{\left(\frac{\lambda_2}{\lambda_1}\right)^k}v_{0_2}p_2+\cancel{\sum_{i=3}^{n}\left(\frac{\lambda_i}{\lambda_1}\right)^k v_{0_i}p_i}\right) \\
	&=\lim_{k\to\infty}\lambda_1^k\left(v_{0_1}p_1+v_{0_2}p_2\right)
\end{align*}
 If $\lambda_1\neq\lambda_2$, $\lim_{k\to\infty}A^k$ is not an eigenvector of $A$.

@@

Next, a comparison of the time taken. The Power Method is so drastically different from the $QR$ Algorithms we will only compare the Basic and Improved Algorithms. Since the Francis double-step algorithm is too complicated, we will modify (``un-improve'') the Improved Algorithm by performing a Hessenberg reduction, then applying the Basic $QR$ Algorithm to the Hessenberg matrix.

We will perform the following procedure for a constant $m$ number of iterations:
\begin{enumerate}
	\item Set $A$ to some random $n\times n$ matrix.
	\item Calculate the time to run $m$ iterations of the Basic $QR$ Algorithm on $A$.
	\item Calculate the time to calculate the Hessenberg matrix $H$ from $A$, and to run $m$ iterations of the Basic $QR$ Algorithm on $H$.
	\item Repeat steps 2 and 3 some number of times, then take the average of each elapsed time.
	\item Increment $n$ by some amount, then start back from step 1.
\end{enumerate}

\begin{center}
	\resizebox{0.49\textwidth}{!}{\input{80.pgf}}
	\resizebox{0.49\textwidth}{!}{\input{120.pgf}}

	\resizebox{0.49\textwidth}{!}{\input{160.pgf}}
	\resizebox{0.49\textwidth}{!}{\input{200.pgf}}

	Time for $m$ iterations of both $QR$ Algorithms for different matrix sizes.
\end{center}

As seen above, the Hessenberg $QR$ Algorithm is indeed faster for larger matrices and iteration counts. At lower iterations, the cost of Hessenberg reduction is higher than the savings of having a sparser matrix.
